---
title: "asdf"
author: "Ingvild, Melissa, Jan"
date: "2023-09-15"
output:
  html_document:
    toc: true
    toc_depth: 2
---

##Read and tidy the dataset
Datasets contains multiple entries for some partient ids. Therefore we use `distinct` to remove these and overwrite the Dataframes `data1` and `data2`.
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(skimr)
library(patchwork)
library(devtools)
library(ggplot2)
library(usethis)
data1 <- read_delim(here("data_original", "exam_data.txt"))
data1 <- distinct(data1) # removes duplicated columns, always check
data2 <- read_delim(here("data_original", "exam_data_join.txt"))
data2 <- distinct(data2)
```

To check the overlap of both dataframes `antijoin` was :
```{r}
antijoined <- anti_join(data1, data2) #no not matchings
```
No entries are generated using `antijoin`, which means that patient IDs from both dataframes match perfectly.
Therefore IDs and can be joined via `joinleft`.

```{r}
merged_data <- left_join(data1, data2)
```
Unwanted columns like year, month and preOp_gender (2 gender columns with identical entries) were removed.
```{r}
merged_data <- subset(merged_data, select = -c(month, year, preOp_gender))
```
The column `preOp_ASA_Mallampati` contained two entries and was split into two columns `asa` and `mallampati`. For better understanding columnheads were renamed into `bmi`, `age`, `smoking`, `pain` and `treatment`.
```{r}
merged_data2 <- merged_data %>%
  separate(col = preOp_ASA_Mallampati, 
           into = c("asa", "mallampati"), 
           sep = "_") %>% # splits column into two
  rename(gender = '1gender',
         bmi = `BMI kg/m2`,
         age = preOp_age,
         smoking = preOp_smoking,
         pain = preOp_pain,
         treatment = treat
         ) # renames column head
```
For better understanding the gender code for male and female patients was changed from 0 and 1 to "Male" and "Female". Same is true for smoking, pain and treatment columns. Using `separate` two separate a `chr` column into two, generates two `chr` columns. Converting `asa` and `mallampati` into numeric entries using `as.numeric`.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(gender = case_when(gender == "0" ~ "Male",
                            gender == "1" ~ "Female",
                            TRUE ~"Other")) %>% #gender to "Male" and "Female" instead of "0"/"1"
  mutate(smoking = case_when(smoking == 1 ~ "Current",
                             smoking == 2 ~ "Past",
                             smoking == 3 ~ "Never")) %>%
  mutate(pain = case_when(pain == 0 ~ "No",
                          pain == 1 ~ "Yes")) %>%
  mutate(treatment = case_when(treatment == 0 ~ "No",
                               treatment == 1 ~ "Yes")) %>%
  mutate(asa = as.numeric(asa),
         mallampati = as.numeric(mallampati)) #converts characters into numeric, very useful
```
Created new sets of columns: `changeincough` and `changeinpain` showing if severity of cough bzw. throat pain changed. 
In 'bmi_4groups' BMI is being cut into four groups of equal size.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(extubation_cough=as.factor(extubation_cough)) %>% #glimpse()
  mutate(changeincough = if_else(pod1am_cough == extubation_cough, "No", "Yes")) %>% #count(changeincough) task a column showing whether severity of cough changed from "extubation" to "pod1am"
  mutate(changeinpain = if_else(pod1am_throatPain == pacu30min_throatPain, "No", "Yes")) %>% #column showing whether severity of throat pain changed from "pacu30min" to "pod1am"
  mutate(bmi_4groups= cut(bmi, 4, labels = c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"))) ## a column cutting BMI into quartiles (4 equal parts); HINT: cut() function
```
Rearranging column sequence using `aragne` function
```{r}
merged_data2 <- merged_data2 %>%
  select(patient_id, bmi, age, smoking, gender, everything()) %>% 
  arrange(patient_id) #- Arrange patient_id column of your dataset in order of increasing number or #alphabetically
```


##Read and tidy the dataset
We stratified our data by a categorical column, 'gender' ('r (merged_data2$gender)') and report min, max, mean and sd of a numeric column, 'age'.
    - Then we stratified the data by a categorical column (only persons with BMI <25, females, persons >50 years of age, or persons who had coughing at extubation) and reported min, max, mean and sd of a numeric column for a defined set of observations. We used age as numeric variable.
        
    - Use two categorical columns in your dataset to create a table (hint: ?count)




The `gender` column includes `r unique(data1$gender)`

The data includes number of co ....... per year from `r min(data1$year)` to `r max(data1$year)`
```{r}
#  Only for persons with BMI <25
merged_data2 %>% 
  filter(bmi<25) %>% 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T))
#Only females
merged_data2 %>% 
  filter(gender== "Female") %>% #only women
  summarise(max(age, na.rm = T), min(age, na.rm = T), mean(age, na.rm = T), sd(age, na.rm=T))
#- Only for persons older than 50 years of age
merged_data2 %>% 
  filter(age> 50) %>% 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T)) 
#Only for persons who had experienced coughing at extubation
merged_data2 %>% 
  filter(as.numeric(extubation_cough) >1) %>% #patient had cough 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T))

```
### Plots showing the distribution of age according to gender.
The distribution seems quite similar between genders in this study

```{r}
plot_gender_age<-ggplot(merged_data2,
                     aes(
                       x=gender,
                       y= age)) +
  geom_jitter(aes(col=gender))
plot_gender_age

boxplot_gender_age2<-ggplot(merged_data2,
                        aes(
                          x=gender,
                          y= age)) +
geom_boxplot(aes(col=gender))
boxplot_gender_age2
# 

#Showing the plots side by side
plot_gender_age + boxplot_gender_age2
```



##Session Info
```{r}
sessionInfo()
```





