---
title: "asdf"
author: "Ingvild, Melissa, Jan"
date: "2023-09-15"
output:
  html_document:
    toc: true
    toc_depth: 2
---

##Read and tidy the dataset
Datasets contains multiple entries for some partient ids. Therefore we use `distinct` to remove these and overwrite the Dataframes `data1` and `data2`.
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(skimr)
library(patchwork)
library(devtools)
library(ggplot2)
library(usethis)
data1 <- read_delim(here("data_original", "exam_data.txt"))
data1 <- distinct(data1) # removes duplicated columns, always check
data2 <- read_delim(here("data_original", "exam_data_join.txt"))
data2 <- distinct(data2)
```

To check the overlap of both dataframes `antijoin` was :
```{r}
antijoined <- anti_join(data1, data2) #no not matchings
```
No entries are generated using `antijoin`, which means that patient IDs from both dataframes match perfectly.
Therefore IDs and can be joined via `joinleft`.

```{r}
merged_data <- left_join(data1, data2)
```
Unwanted columns like year, month and preOp_gender (2 gender columns with identical entries) were removed.
```{r}
merged_data <- subset(merged_data, select = -c(month, year, preOp_gender))
```
The column `preOp_ASA_Mallampati` contained two entries and was split into two columns `asa` and `mallampati`. For better understanding columnheads were renamed into `bmi`, `age`, `smoking`, `pain` and `treatment`.
```{r}
merged_data2 <- merged_data %>%
  separate(col = preOp_ASA_Mallampati, 
           into = c("asa", "mallampati"), 
           sep = "_") %>% # splits column into two
  rename(gender = '1gender',
         bmi = `BMI kg/m2`,
         age = preOp_age,
         smoking = preOp_smoking,
         pain = preOp_pain,
         treatment = treat
         ) # renames column head
```
For better understanding the gender code for male and female patients was changed from 0 and 1 to "Male" and "Female". Same is true for smoking, pain and treatment columns. Using `separate` two separate a `chr` column into two, generates two `chr` columns. Converting `asa` and `mallampati` into numeric entries using `as.numeric`.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(gender = case_when(gender == "0" ~ "Male",
                            gender == "1" ~ "Female",
                            TRUE ~"Other")) %>% #gender to "Male" and "Female" instead of "0"/"1"
  mutate(smoking = case_when(smoking == 1 ~ "Current",
                             smoking == 2 ~ "Past",
                             smoking == 3 ~ "Never")) %>%
  mutate(pain = case_when(pain == 0 ~ "No",
                          pain == 1 ~ "Yes")) %>%
  mutate(treatment = case_when(treatment == 0 ~ "No",
                               treatment == 1 ~ "Yes")) %>%
  mutate(asa = as.numeric(asa),
         mallampati = as.numeric(mallampati)) #converts characters into numeric, very useful
```
Created new sets of columns: `changeincough` and `changeinpain` showing if severity of cough bzw. throat pain changed. 
In 'bmi_4groups' BMI is being cut into four groups of equal size.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(extubation_cough=as.factor(extubation_cough)) %>% #glimpse()
  mutate(changeincough = if_else(pod1am_cough == extubation_cough, "No", "Yes")) %>% #count(changeincough) task a column showing whether severity of cough changed from "extubation" to "pod1am"
  mutate(changeinpain = if_else(pod1am_throatPain == pacu30min_throatPain, "No", "Yes")) %>% #column showing whether severity of throat pain changed from "pacu30min" to "pod1am"
  mutate(bmi_4groups= cut(bmi, 4, labels = c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"))) ## a column cutting BMI into quartiles (4 equal parts); HINT: cut() function
```
Rearranging column sequence using `aragne` function
```{r}
merged_data2 <- merged_data2 %>%
  select(patient_id, bmi, age, smoking, gender, everything()) %>% 
  arrange(patient_id) #- Arrange patient_id column of your dataset in order of increasing number or #alphabetically
```


## tidy the dataset and exploring
# Explore the data 

```{r}
summary (merged_data2)
glimpse (merged_data2)

library(skimr)

skim(merged_data2)

naniar::gg_miss_var(merged_data2) # There is to missing data for several variables, 236 patient were enrolled , we have 235 observations? 

naniar::gg_miss_var(merged_data2, facet = gender) # the missing variables are in menn

```

# Two categorical columns to create a table

```{r}
Table<- merged_data2 %>%
  count(gender, pacu30min_cough)
Table

```

## Plots

```{r}
ggplot(data = merged_data2) +
  aes(
    x = pacu30min_throatPain,
    y = pacu30min_cough
  ) +
  geom_point() +
  geom_smooth(method = "lm")

# There seems to be a correlation between 30 min throatpain and 30 min cough
```

```{r}
ggplot(merged_data2, aes(x=as.factor(treatment), y=age)) +
  geom_col(aes(fill = age), position=position_dodge())

# Yes the age distribution depends on treatment, alot of young people were not treated.  
```


## Statistical test 

```{r}
Test_gender_patient <- merged_data2 %>%
  mutate (treatment=ifelse(treatment=="Yes", 1, 0), 
          gender=ifelse(gender=="Male", 0, 1)) %>% 
  t.test( treatment~gender, data = .)

# Welch Two Sample t-test gives the p-values 0.5, so the treatment does not depend on the patient.
```


```{r}
head(data1)
```


##Session Info
```{r}
sessionInfo()
```





