---
title: "Licorice gargle - A group project during RMED901"
author: "Ingvild, Melissa, Jan"
date: "2023-09-15"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 2
    code_folding: hide
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(skimr)
library(patchwork)
library(devtools)
library(ggplot2)
library(usethis)
```
# Read, tidy & explore
Before analyzing, the data must be read in correctly, cleaned and checked for e.g. missing entries.

## Reading the dataframe
Both provided `.txt` datasets contain headers and entries are separated via `space` or `\t` as delimiter. Therefore they can be read via `read_delim` function, which automatically recognizes the delimiter (but could be specified e.g. via `delim = "\t"`).
Since both datasets contain multiple entries for some partient ids, we use `distinct` to remove duplicates and overwrite the dataframes `data1` and `data2` to generate data without duplicates.
```{r} 
data1 <- read_delim(here("data_original", "exam_data.txt"))
data1 <- distinct(data1) # removes duplicated columns, always check
data2 <- read_delim(here("data_original", "exam_data_join.txt"))
data2 <- distinct(data2)
```

## Merging dataframes
To check the overlap of both dataframes `antijoin` was used. No entries are generated using `antijoin`, which means that patient IDs from both dataframes match perfectly.
```{r}
antijoined <- anti_join(data1, data2) #no not matchings
```

Therefore IDs and can be joined using e.g. `joinleft`.

```{r}
merged_data <- left_join(data1, data2)
```

## Tidying dataframe
Unwanted columns like `year`, `month` and `preOp_gender` (`merged_data` datagrame cointains 2 gender columns with identical entries) were removed.
```{r}
merged_data <- subset(merged_data, select = -c(month, year, preOp_gender))
```
The column `preOp_ASA_Mallampati` contained two entries and was split into two columns `asa` and `mallampati`. For better understanding columnheads were renamed into `bmi`, `age`, `smoking`, `pain` and `treatment`.
```{r}
merged_data2 <- merged_data %>%
  separate(col = preOp_ASA_Mallampati, 
           into = c("asa", "mallampati"), 
           sep = "_") %>% # splits column into two
  rename(gender = '1gender',
         bmi = `BMI kg/m2`,
         age = preOp_age,
         smoking = preOp_smoking,
         pain = preOp_pain,
         treatment = treat
         ) # renames column head
```
To improve readability, the initially used gender code for male and female patients was changed from `0` and `1` to `"Male"` and `"Female"`. Same is true for `smoking`, `pain` and `treatment` columns. Using `separate` two separate a `chr` column into two, generates two `chr` columns. Converting `asa` and `mallampati` into numeric entries using `as.numeric`.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(gender = case_when(gender == "0" ~ "Male",
                            gender == "1" ~ "Female",
                            TRUE ~"Other")) %>% #gender to "Male" and "Female" instead of "0"/"1"
  mutate(smoking = case_when(smoking == 1 ~ "Current",
                             smoking == 2 ~ "Past",
                             smoking == 3 ~ "Never")) %>%
  mutate(pain = case_when(pain == 0 ~ "No",
                          pain == 1 ~ "Yes")) %>%
  mutate(treatment = case_when(treatment == 0 ~ "No",
                               treatment == 1 ~ "Yes")) %>%
  mutate(asa = as.numeric(asa),
         mallampati = as.numeric(mallampati)) #converts characters into numeric, very useful
```
The next part of the script creates two new columns "changeincough" and "changeinpain" that indicate whether the severity of the cough resp. throat pain has changed.
In 'bmi_4groups' BMI is being cut into four groups of equal size.
```{r}
merged_data2 <- merged_data2 %>%
  mutate(extubation_cough=as.factor(extubation_cough)) %>% #glimpse()
  mutate(changeincough = if_else(pod1am_cough == extubation_cough, "No", "Yes")) %>% #count(changeincough) task a column showing whether severity of cough changed from "extubation" to "pod1am"
  mutate(changeinpain = if_else(pod1am_throatPain == pacu30min_throatPain, "No", "Yes")) %>% #column showing whether severity of throat pain changed from "pacu30min" to "pod1am"
  mutate(bmi_4groups = cut(bmi, 4, labels = c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"))) ## a column cutting BMI into quartiles (4 equal parts); HINT: cut() function
```
Rearranging column sequence using `arrange` function.
```{r}
merged_data2 <- merged_data2 %>%
  select(patient_id, bmi, age, smoking, gender, everything()) %>% 
  arrange(patient_id) #- Arrange patient_id column of your dataset in order of increasing number or #alphabetically
```

## Exploring
The merged dataset contains 235 different patients (rows) and 23 different columns.
```{r}
glimpse (merged_data2)
```

Next we screened for missing variables in `"Male"`.
```{r}
#summary (merged_data2)
#glimpse (merged_data2)
#skim(merged_data2)
#naniar::gg_miss_var(merged_data2) # There is to missing data for several variables, 236 patient were enrolled , we have 235 observations? 
naniar::gg_miss_var(merged_data2, facet = gender) # the missing variables are in menn
```

Patients with the ID's 113 and 123 had no data entries in the columns with the missing entries.
```{r}
which(is.na(merged_data2),arr.ind=TRUE)
```

We stratified our data by a categorical column, `gender` (r (merged_data2$gender)) and report min, max, mean and sd of a numeric column, `age`. Then we stratified the data by a categorical column containing only persons with BMI <25,
```{r}
#  Only for persons with BMI <25
merged_data2 %>% 
  filter(bmi<25) %>% 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T))
```
females
```{r}
#Only females
merged_data2 %>% 
  filter(gender== "Female") %>% #only women
  summarise(max(age, na.rm = T), min(age, na.rm = T), mean(age, na.rm = T), sd(age, na.rm=T))
```
patients older than 50 years
```{r}
#- Only for persons older than 50 years of age
merged_data2 %>% 
  filter(age> 50) %>% 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T))
```
and persons who had coughing at extubation.
```{r}
#Only for persons who had experienced coughing at extubation
merged_data2 %>% 
  filter(as.numeric(extubation_cough) >1) %>% #patient had cough 
  summarise(max(age, na.rm = T), 
            min(age, na.rm = T), 
            mean(age, na.rm = T), 
            sd(age, na.rm=T))

```

## Two categorical columns to create a table

```{r}
Table<- merged_data2 %>%
  count(gender, smoking)
knitr::kable(Table)
```


# Plots
Plot showing correlation between throat pain and cough 30 min after treatment.
```{r}
ggplot(data = merged_data2) +
  aes(
    x = pacu30min_throatPain,
    y = pacu30min_cough
  ) +
  geom_point() +
  geom_smooth(method = "lm")

# There seems to be a correlation between 30 min throatpain and 30 min cough
```

Plot showing if age depends on treatment.
```{r}
ggplot(merged_data2, aes(x=as.factor(treatment), y=age)) +
  geom_col(aes(fill = age), position=position_dodge())

# Yes the age distribution depends on treatment, alot of young people were not treated.  
```
Plots showing the distribution of age according to gender.
The distribution seems quite similar between genders in this study


```{r, echo = FALSE}
plot_gender_age<-ggplot(merged_data2,
                     aes(
                       x=gender,
                       y= age)) +
  geom_jitter(aes(col=gender))

boxplot_gender_age2<-ggplot(merged_data2,
                        aes(
                          x=gender,
                          y= age)) +
geom_boxplot(aes(col=gender))

plot_gender_age + boxplot_gender_age2

```

# Statistical tests 
Is the treatment dependent on the patient's gender? Before we can perform a test we have to mutate the `chr` entries in `gender` back to `num` entries using `mutate`. To answer this question we performed a *Welch's t-test* using `t.test` function and neatly display the results using `broom::tidy`. The resulting p-value > 0.5 indicates no dependency of th treatment to the patient's gender.
```{r}
#Does the treatment depend on the patient's gender
Test_gender_vs_patient <- merged_data2 %>%
  mutate (treatment=ifelse(treatment=="Yes", 1, 0), 
          gender=ifelse(gender=="Male", 0, 1)) %>% 
  t.test( treatment~gender, data = .)
  broom::tidy(Test_gender_vs_patient)
```

Next, we investigated whether treatment might depend on patients smoking using a *Welch's t-test*. Again, we had to change the `chr` entries in `treatment` and `smoking` back to `num` entries. The resulting p-value > 0.9 reveals no treatment difference between smokers and non-smokers. 
```{r}
#Does the treatment depend on the preoperative smoking
Test_treatment_vs_smoking <- merged_data2 %>%
  mutate(treatment=ifelse(treatment=="Yes", 1, 0),
         smoking=ifelse(smoking=="Never", 0, 1)) %>% 
  t.test(treatment~smoking, data = .)
  broom::tidy(Test_treatment_vs_smoking)
```
Lastly, we checked if there is a dependency between treatment and preoperative pain. Here we found a strong correlation between treatment and preoperative pain.
```{r}
#Does the treatment depend on the preoperative pain
Test_treatment_vs_pain <- merged_data2 %>%
  mutate(treatment=ifelse(treatment=="Yes", 1, 0), 
         pain=ifelse(pain=="Yes", 1, 0)) %>% 
  t.test(treatment~pain, data = .)
broom::tidy(Test_treatment_vs_pain)
```

# Session Info
```{r}
sessionInfo()
```