---
title: "Group 2"
author: "Ingvild, Melissa, Jan"
date: "2023-09-15"
output: html_document:
  toc: true
  code_folding: hide 
---

##Read and tidy the dataset
```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(skimr)
library(patchwork)
library(devtools)
library(ggplot2)
library(usethis)
data1 <- read_delim(here("../data_original", "exam_data.txt"))
data2 <- read_delim(here("../data_original", "exam_data_join.txt"))
```
Datasets contains multiple entries for some partient ids. Therefore we use `distinct` to remove these and overwrite the Dataframes `data1` and `data2`. Now both dataframes contain the same patient IDs and can be joined via `joinleft`.

```{r setup, include=FALSE} 
data1 <- distinct(data1) # removes duplicated columns, always check
data2 <- distinct(data2)
```



The `gender` column includes `r unique(data1$gender)`

The data includes number of co ....... per year from `r min(data1$year)` to `r max(data1$year)`
```{r}
head(data1)
```


##Session Info
```{r}
sessionInfo()
```





